---
title: "Mutational signatures extraction"
author: "Romina Royo & Ferran Nadeu"
date: "`r format(Sys.time(), '%d/%m/%y')`"
output:
  html_document: 
    theme: paper
    toc: yes
    toc_float: true
    toc_depth: 3
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

# Settings

## Load libraries

```{r packages,message=FALSE}
library("BSgenome")
ref_genome <- "BSgenome.Hsapiens.UCSC.hg19"
library(ref_genome, character.only = TRUE)
library("ggplot2")
library("ggrepel")
library("gridExtra")
library("ComplexHeatmap")
library("circlize")
library(data.table)
library(MutationalPatterns)
library(nnls)
library(lattice)
library(lsa)
library(openxlsx)
# Custom functions
source("mut_sigs_functions.R")
```


## Load data

```{r mutmat, cache=TRUE, fig.align='center'}
# Results
hdp_all <- fread("supplementary_tables/supplementaryTableX_hdp_extracted_signatures.tsv",header=TRUE)
sigpro_all <- fread("supplementary_tables/supplementaryTableX_sigpro_extracted_signatures.tsv",header=TRUE)
siganal_all <- fread("supplementary_tables/supplementaryTableX_siganal_extracted_signatures.tsv",header=TRUE)
sigfit_all <- fread("supplementary_tables/supplementaryTableX_sigfit_extracted_signatures.tsv",header=TRUE)
hdp <- hdp_all
sigpro <- sigpro_all
siganal <- siganal_all
sigfit <- sigfit_all

# Mut types order
mut_types <- fread("additional_data/mutTypes_order.txt",header=FALSE)

# COSMIC
cancer_signatures <- fread("additional_data/COSMIC_v3.2_SBS_GRCh37.txt", header = T, stringsAsFactors = F)
cancer_signatures <- cancer_signatures[match(mut_types$V1,cancer_signatures$Type),]
cancer_signatures <- cancer_signatures[,-c(1)]
cancer_signatures <- as.data.frame(cancer_signatures)

# SBSA from Kanter et al.
SBSA <- read.xlsx("additional_data/mmc3-2.xlsx")
colnames(SBSA) <- c("context","SBSA")
SBSA$SBSA <- as.numeric(SBSA$SBSA)
# SBS-MM1 from Rustad et al.
load(file="additional_data/signature_ref.rda")
SBSMM1 <- signature_ref[,c("SBS-MM1"),drop=FALSE]
# SBS signatures from Kucab et al.
kucab <- fread("additional_data/Mutagen53_sub_signature.txt")
kucab <- kucab[,-1]
kucab <- as.data.frame(kucab)
```


# Plot extracted signatures

Plot extracted signatures by each program: hdp, sigprofiler, signatureAnalyzer, sigfit

```{r,fig.width=10}
# Sigprofiler
p <- plot_extracted_sigs(sigpro,mut_types$V1,"SigProfiler")
p
pdf("outputs/sigprofiler_extracted_signatures.pdf",width = 10,height = 5.8,useDingbats = F)
p
dev.off()

# Hdp
# Plot potential artefacts
artefacts <- paste0("hdp_",seq(11,ncol(hdp),1))
hdp <- as.data.frame(hdp)
p <- plot_extracted_sigs(hdp[,colnames(hdp) %in% artefacts],mut_types$V1,"HDP")
pdf("outputs/hdp_extracted_signatures_potential_artefacts.pdf",width = 10,height = 5.8,useDingbats = F)
p
dev.off()
# Remove potential artefacts
hdp <- hdp[,!colnames(hdp) %in% artefacts]
p <- plot_extracted_sigs(hdp,mut_types$V1,"HDP")
p
pdf("outputs/hdp_extracted_signatures.pdf",width = 10,height = 12,useDingbats = F)
p
dev.off()

# SignatureAnalyzer
p <- plot_extracted_sigs(siganal,mut_types$V1,"SignatureAnalyzer")
p
pdf("outputs/SignatureAnalyzer_extracted_signatures.pdf",width = 10,height = 8,useDingbats = F)
p
dev.off()

# Sigfit
p <- plot_extracted_sigs(sigfit,mut_types$V1,"sigfit")
p
pdf("outputs/sigfit_extracted_signatures.pdf",width = 10,height = 5.8,useDingbats = F)
p
dev.off()
```


# Assign and decompose extracted signatures

We assing or deconvolute all extracted signatures into known signatures from COSMIC/PCAWG v3.2 catalogue
We also compared novel signatures to the compendium of mutational signatures of environmental agents from Kucab et al., and a novel signature from Kanter et al.
 
## Initial considerations

```{r,fig.height=2,fig.width=10}
# SBS5 composition from hdp extracted signatures: SBS5 was directly extracted by all algorithms except for hdp. However, we identified two signatures that actually compose SBS5
mod1 <- nnls(as.matrix(hdp[,c("hdp_2","hdp_3")]),cancer_signatures[,c("SBS5")])
mergesig <- mergeSignature2(as.data.frame(hdp),c("hdp_2","hdp_3"),mod1$x,mut_types$V1,"SBS5=hdp_2+hdp_3")
mergesig[[2]]
weights <- paste(round(mod1$x,2),collapse="_")
weights_norm <- paste(round(mod1$x/sum(mod1$x),2),collapse="_")
cossim <- round(cosineSimilarity(mergesig[[1]],cancer_signatures[,c("SBS5")]),2)
print(paste0("hdp2+hdp3 - SBS5 ",weights_norm," ",cossim))

# SBS2 and SBS13 APOBEC-related signatures can be visually recognized in one sample (839-01-01BD). By visual inspection of the extracted signatures, we identified their extraction by hdp and signatureAnalyzer.

## Sample exhibiting APOBEC signatures
muts <- fread("supplementary_tables/supplementaryTableX_all_wgs_mutations.tsv")
muts <- muts[muts$TYPE=="SNV",]
vcfs2 <- makeGRangesFromDataFrame(muts, keep.extra.columns=TRUE, ignore.strand=TRUE,
                         seqinfo=NULL, seqnames.field="CHROM", start.field="POSITION",
                         end.field="POSITION", starts.in.df.are.0based=FALSE)
vcfs2 <- split(vcfs2, as.factor(vcfs2$SAMPLE))
seqlevelsStyle(vcfs2) <- "UCSC"
genome(vcfs2) <- "hg19"
mut_mat <- mut_matrix(vcf_list = vcfs2, ref_genome = ref_genome)
p <- plotSignature2(mut_mat[,c("839-04-01BD")],"839-04-01BD with APOBEC",mut_types$V1)
pdf("outputs/839-APOBEC.pdf", height=3, width=10,useDingbats = F)
p
dev.off()

## HDP
mod1 <- nnls(as.matrix(cancer_signatures[,c("SBS2","SBS13")]),hdp$hdp_7)
mergesig <- mergeSignature2(as.data.frame(cancer_signatures),c("SBS2","SBS13"),mod1$x,mut_types$V1,"hdp_7=SBS2+SBS13")
mergesig[[2]]
weights <- paste(round(mod1$x,2),collapse="_")
weights_norm <- paste(round(mod1$x/sum(mod1$x),2),collapse="_")
cossim <- round(cosineSimilarity(mergesig[[1]],hdp$hdp_7),2)
print(paste0("SBS2 + SBS13 - hdp_7 ",weights_norm," ",cossim))
hdp_assigned <- c("hdp_7")

## SignatureAnalyzer
mod1 <- nnls(as.matrix(cancer_signatures[,c("SBS2","SBS13","SBS5")]),siganal$siganal_5)
mergesig <- mergeSignature2(as.data.frame(cancer_signatures),c("SBS2","SBS13","SBS5"),mod1$x,mut_types$V1,"siganal_5=SBS2+SBS13+SBS5")
mergesig[[2]]
weights <- paste(round(mod1$x,2),collapse="_")
weights_norm <- paste(round(mod1$x/sum(mod1$x),2),collapse="_")
cossim <- round(cosineSimilarity(mergesig[[1]],siganal$siganal_5),2)
print(paste0("SBS2 + SBS13 + SBS5 - siganal_5 ",weights_norm," ",cossim))

# Recently published SBSA from Kanter et al. corresponds to hdp_9 and siganal_7
c <- cosineSimilarity(hdp$hdp_9,as.numeric(SBSA$SBSA))
print(paste0("SBSA - hdp_9, cosine similarity: ",c))
c <- cosineSimilarity(siganal$siganal_7,as.numeric(SBSA$SBSA))
print(paste0("SBSA - siganal_7, cosine similarity: ",c))

# Already assigned/decomposed signatures
hdp_assigned <- c(hdp_assigned,"hdp_2","hdp_3","hdp_9")
siganal_assigned <- c("siganal_5","siganal_7")


# Save SBS-ganciclovir and SBS-RT (see below) to be used by other scripts
sbsrt <- hdp$hdp_9
sbsrt <- as.data.frame(sbsrt)
sbsrt$MutationType <- mut_types$V1
sbsrt <- sbsrt[,c(2,1)]
colnames(sbsrt) <- c("MutationType","SBS-ganciclovir")
sbsrt$`SBS-RT` <- hdp$hdp_10
write.table(sbsrt,"outputs/SBS-ganciclovir_RT2.tsv", sep="\t", col.names=T, row.names = F, quote=F)
```

## PCAWG similarities, ONE signature

We compare all extracted signatures to the COSMIC catalogue. If the cosine similarity is greater than 0.85 we assign the corresponding signature.

```{r,fig.width=14}
assign_to_one_pcawg <- function(signatures,cancer_signatures){
# Create a table that summarize all cosine similarities for our extracted signatures
signatures <- as.data.frame(signatures)
f <- data.frame(matrix(0, ncol=ncol(cancer_signatures), nrow = ncol(signatures)))
colnames(f) <- colnames(cancer_signatures)
rownames(f) <- colnames(signatures)
for (i in 1:ncol(signatures)) {
  col_sample <- signatures[,i]
  for (j in 1:ncol(cancer_signatures)) {
    col_cosmic <- cancer_signatures[,j]
    c <- cosineSimilarity(col_cosmic,col_sample)
    f[i,j] <- c
  }
}

f[] <- lapply(f,function(x)round(x,2))
to_rm <- which(is.na(apply(f, 2, function(x) ifelse(max(x, na.rm = TRUE)>0.2,x,NA))))
f <- f[,-to_rm]

h <- Heatmap(f, col = colorRamp2(c(0, 0.6,1), c("white","white", "red")), rect_gp = gpar(col = "black", lty = 1, lwd = 1), 
             cluster_rows = FALSE, cluster_columns = FALSE, name = "Cosine",
             heatmap_legend_param = list(color_bar = "continuous", at = c(0, 0.6,1), labels = c("0","0.6", "1"), legend_height = unit(4, "cm")))

max_cossim <- apply(f,1,function(x) which(x==max(x)))

max_cossim_df <- data.frame()
for (i in 1:nrow(f)){
  aux <- data.frame(sig=rownames(f)[i],cosmic_sig=paste(colnames(f)[max_cossim[[i]]],collapse="_OR_"),cossim=paste(f[i,max_cossim[[i]]],collapse="_OR_"))
  max_cossim_df <- rbind(max_cossim_df,aux)
}

max_cossim_df

return(list(h,max_cossim_df))
}

# Remove already decomposed sigs
siganal <- as.data.frame(siganal)
rownames(siganal) <- mut_types$V1
siganal <- siganal[,!colnames(siganal) %in% siganal_assigned]

hdp <- as.data.frame(hdp)
rownames(hdp) <- mut_types$V1
hdp <- hdp[,!colnames(hdp) %in% hdp_assigned]
```

### SignatureAnalyzer

```{r,fig.width=18,fig.height=5}
res <- assign_to_one_pcawg(siganal,cancer_signatures)
res[1]
res[2]

# We assign:
# siganal_1 - SBS1, 0.94
# siganal_2 - SBS5, 0.88
# siganal_3 - SBS9, 0.98
# siganal_4 - SBS18, 0.89
# siganal_6 - SBS17b, 0.94

siganal_assigned <- c("siganal_1","siganal_2","siganal_3","siganal_4","siganal_6")
```

### HDP

```{r,fig.width=18,fig.height=5}
res <- assign_to_one_pcawg(hdp,cancer_signatures)
res[1]
res[2]

# We assign:
# hdp_1 - SBS1, 0.97
# hdp_4 - SBS8, 0.86
# hdp_5 - SBS9, 0.94
# hdp_6 - SBS18, 0.96
# hdp_8 - SBS17b, 0.92

hdp_assigned <- c(hdp_assigned,"hdp_1","hdp_4","hdp_5","hdp_6","hdp_8")
```

### SigProfiler

```{r,fig.width=18,fig.height=4}
res <- assign_to_one_pcawg(sigpro,cancer_signatures)
res[1]
res[2]

# We assign:
# sigpro_2 - SBS5, 0.9
# sigpro_3 - SBS9, 0.99

sigpro_assigned <- c("sigpro_2","sigpro_3")
```

### sigfit

```{r,fig.width=18,fig.height=4}
res <- assign_to_one_pcawg(sigfit,cancer_signatures)
res[1]
res[2]

# We assign:
# sigfit_2 - SBS5, 0.87
# sigfit_3 - SBS9, 0.98

sigfit_assigned <- c("sigfit_2","sigfit_3")
```

## Decompose into N mutational signatures

We decompose the remaining extracted signatures (which could not be previously assigned to one COSMIC mutational signature with a cosine similarity >0.85) into N mutational signatures using an expectation maximization (EM) approach, based on:
Lee-Six, H., Olafsson, S., Ellis, P. et al. The landscape of somatic mutation in normal colorectal epithelial cells. Nature 574, 532–537 (2019). https://doi-org.sire.ub.edu/10.1038/s41586-019-1672-7

First, we try to deconvolute the remaining signatures into the previously identified signatures. If the cosine similarity between the reconstituted signature and the original one is greater than 0.85, we assign its EM deconvolution.

Next, we try to break down the remaining signatures using all PCAWG signatures. To avoid overfitting, we initially consider PCAWG signatures contributing more than 0.1, as suggested by the authors. If the cosine similarity between the reconstituted signature and the original one is below 0.85, we consider it a novel mutational signature.

We finally show how the addition of the novel signature SBS-RT (selected from hdp for the sake of lesser background noise) can deconvolute the corresponding novel signatures extracted by the other programs (SigProfiler and sigfit) more accurately.



```{r}
decompose_into_N_pcawg <- function(signatures,sigs_name,cancer_signatures,pdf_file=NULL){

  sample_list <- colnames(signatures)
  mutations <- signatures[,sample_list]
  mutations <- as.data.frame(mutations)

  cancer_signatures_t <- t(cancer_signatures)
  signatures_names <- rownames(cancer_signatures_t)
  signature_fraction = array(NA,dim=c(dim(cancer_signatures_t)[1], length(sample_list)))
  rownames(signature_fraction) = signatures_names
  colnames(signature_fraction) = sample_list
  num_signatures = length(signatures_names)
  maxiter <- 1000

  for (j in 1:length(sample_list)) {
    mut_freqs = mutations[,j]
    mut_freqs[is.na(mut_freqs)] = 0

    # EM algowith to estimate the signature contribution
    alpha = runif(num_signatures); alpha=alpha/sum(alpha) # Random start (seems to give ~identical results)
    # alpha = rep(1/num_signatures,num_signatures) # Uniform start
    for (iter in 1:maxiter) {
      contr = t(array(alpha,dim=c(num_signatures,96))) * t(cancer_signatures_t)
      probs = contr/array(rowSums(contr),dim=dim(contr))
      probs = probs * mut_freqs
      old_alpha = alpha
      alpha = colSums(probs)/sum(probs)
      if (sum(abs(alpha-old_alpha))<1e-5) {
        break
      }
    }
    # Saving the signature contributions for the sample
    print(j/length(sample_list))
    signature_fraction[,j] = alpha
  }

  s <- signature_fraction
  rownames(s) <- paste0("pcawg_", rownames(s))
  colnames(s) <- paste0(sigs_name,"_", colnames(s))
  
  color.palette = colorRampPalette(c("white", "orange", "purple"))
  if (ncol(signatures)>1){
    # print(levelplot((s[dim(s)[1]:1,]),col.regions=color.palette, aspect="fill", scales=list(x=list(rot=90))))
  }

  signature_fraction_bk <- signature_fraction

  ## Continue...
  # For each signature, select cancer signatures with contribution > 0.1 and reconstruct original signature.
  # Now, we only keep cancer_signatures contributing more than > 0.1
  for (j in 1:length(sample_list)) {
    constit <- rownames(signature_fraction_bk[signature_fraction_bk[,sample_list[j]]>0.1,,drop=FALSE])
    constit
    if (length(constit)>1){
      gdsigs <- constit
      sigs <- as.matrix(cancer_signatures[,gdsigs])
      signatures2 <- t(sigs)

      aa <- sample_list[j]
      mutations <- signatures[,aa,drop=FALSE]
      head(mutations)
      signatures_names <- rownames(signatures2)
      signature_fraction = array(NA,dim=c(dim(signatures2)[1], 1))
      rownames(signature_fraction) = signatures_names
      colnames(signature_fraction) = sample_list[j]
      num_signatures = length(signatures_names)
      maxiter <- 1000

      mut_freqs = mutations[[1]]
      mut_freqs[is.na(mut_freqs)] = 0

      # EM algowith to estimate the signature contribution
      alpha = runif(num_signatures); alpha=alpha/sum(alpha) # Random start (seems to give ~identical results)
      # alpha = rep(1/num_signatures,num_signatures) # Uniform start
      for (iter in 1:maxiter) {
        contr = t(array(alpha,dim=c(num_signatures,96))) * t(signatures2)
        probs = contr/array(rowSums(contr),dim=dim(contr))
        probs = probs * mut_freqs
        old_alpha = alpha
        alpha = colSums(probs)/sum(probs)
        if (sum(abs(alpha-old_alpha))<1e-5) {
          break
        }
      }

      # Saving the signature contributions for the sample
      signature_fraction[,1] = alpha
      print(signature_fraction)

      s <- signature_fraction
      rownames(s) <- paste0("pcawg_", rownames(s))
      colnames(s) <- paste0(sigs_name,"_", colnames(s))
      dim(s)
      reconsbs <- rep(0, 96)
      for (ct in constit) {
        reconsbs <- reconsbs + (cancer_signatures[,ct]*s[paste0("pcawg_", ct),paste0(sigs_name,"_",sample_list[j])])
      }
      reconsbs
      sum(reconsbs)
      cosine(x=reconsbs, y=signatures[[aa]]) # 0.99

      # Plot the original signature broken down into their composite signatures 
      if (!is.null(pdf_file)){
        pdf(paste0(pdf_file,"_",sample_list[j],".pdf"),useDingbats = F)
        par(mfrow=c(6,1))
        par(mar=c(1,2,4,1))
        hist.cols <- rep(c("lightblue", "black", "red", "gray", "lightgreen", "lightpink"), each=16)
        barplot(signatures[[aa]], col=hist.cols, main=paste0(sigs_name,"_",sample_list[j]))
        barplot(reconsbs, col=hist.cols, main=paste0("Reconstituted ",sigs_name," ",sample_list[j],", cosine similarity to original: ", ... = round(cosine(x=reconsbs, y=signatures[[aa]]), digits=2)))
        for (ct in constit) {
          barplot(cancer_signatures[,ct], col=hist.cols, main=paste0("PCAWG ", ct, " accounts for ", round(s[paste0("pcawg_", ct),paste0(sigs_name,"_",sample_list[j])], digits=2)))
        }
        dev.off()
      }
      par(mfrow=c(5,1))
      par(mar=c(1,2,4,1))
      hist.cols <- rep(c("blue", "black", "red", "grey", "green", "pink"), each=16)
      barplot(signatures[[aa]], col=hist.cols, main=paste0(sigs_name,"_",sample_list[j]))
      barplot(reconsbs, col=hist.cols, main=paste0("Reconstituted ",sigs_name," ",sample_list[j],", cosine similarity to original: ", ... = round(cosine(x=reconsbs, y=signatures[[aa]]), digits=2)))
      for (ct in constit) {
        barplot(cancer_signatures[,ct], col=hist.cols, main=paste0("PCAWG ", ct, " accounts for ", round(s[paste0("pcawg_", ct),paste0(sigs_name,"_",sample_list[j])], digits=2)))
      }
    } else {
      print(paste0("Not enough sigs ",sigs_name,"_",sample_list[j]))
    }
  }
}

# Remove previously decomposed signatures
siganal <- siganal[,!colnames(siganal) %in% siganal_assigned,drop=FALSE]

hdp <- hdp[,!colnames(hdp) %in% hdp_assigned,drop=FALSE]

sigpro <- as.data.frame(sigpro)
sigpro <- sigpro[,!colnames(sigpro) %in% sigpro_assigned,drop=FALSE]

sigfit <- as.data.frame(sigfit)
sigfit <- sigfit[,!colnames(sigfit) %in% sigfit_assigned,drop=FALSE]

# Adding SBS-ganciclovir and SBS-RT to the catalogue for future use
cancer_signatures2 <- cbind(cancer_signatures,sbsrt[,c(2,3),drop=FALSE])
```


### SignatureAnalyzer

```{r}
# Using previously identified signatures
decompose_into_N_pcawg(siganal,"siganal",cancer_signatures2[,c("SBS1","SBS2","SBS5","SBS8","SBS13","SBS18","SBS9","SBS17b","SBS-ganciclovir")])

# Using all PCAWG signatures
decompose_into_N_pcawg(siganal,"siganal",cancer_signatures)

# We already saw that siganal_8 has a cosine similarity of 0.941 with SBS-RT
```

### HDP

```{r}
# Using previously identified signatures
decompose_into_N_pcawg(hdp,"hdp",cancer_signatures2[,c("SBS1","SBS2","SBS5","SBS8","SBS13","SBS18","SBS9","SBS17b","SBS-ganciclovir")])

# Using all PCAWG signatures
decompose_into_N_pcawg(hdp,"hdp",cancer_signatures,"outputs/SBS-RT_decompose")

# hdp_10 is considered the novel mutational process: SBS-RT
```


### SigProfiler

```{r}
# Using previously identified signatures
decompose_into_N_pcawg(sigpro,"sigpro",cancer_signatures2[,c("SBS1","SBS2","SBS5","SBS8","SBS13","SBS18","SBS9","SBS17b","SBS-ganciclovir")])

# We assign:
# sigpro_1 - SBS1 + SBS5, 0.98
# sigpro_4 - SBS5 + SBS8 + SBS18, 0.9

sigpro_assigned <- c(sigpro_assigned,"sigpro_1","sigpro_4")
sigpro <- sigpro[,!colnames(sigpro) %in% sigpro_assigned,drop=FALSE]

# Using all PCAWG signatures
decompose_into_N_pcawg(sigpro,"sigpro",cancer_signatures)

# Using all PCAWG signatures + novel SBS-RT
decompose_into_N_pcawg(sigpro,"sigpro",cancer_signatures2[,c("SBS1","SBS2","SBS5","SBS8","SBS13","SBS18","SBS9","SBS17b","SBS-ganciclovir","SBS-RT")])
```

### sigfit

```{r}
# Using previously identified signatures
decompose_into_N_pcawg(sigfit,"sigfit",cancer_signatures2[,c("SBS1","SBS2","SBS5","SBS8","SBS13","SBS18","SBS9","SBS17b","SBS-ganciclovir")])

# We assign:
# sigfit_1 - SBS1 + SBS5, 0.98
# sigfit_4 - SBS5 + SBS8 + SBS18, 0.91

sigfit_assigned <- c(sigfit_assigned,"sigfit_1","sigfit_4")
sigfit <- sigfit[,!colnames(sigfit) %in% sigfit_assigned,drop=FALSE]

# Using all PCAWG signatures
decompose_into_N_pcawg(sigfit,"sigfit",cancer_signatures)

# Using all PCAWG signatures + novel SBS-RT
decompose_into_N_pcawg(sigfit,"sigfit",cancer_signatures2[,c("SBS1","SBS2","SBS5","SBS8","SBS13","SBS18","SBS9","SBS17b","SBS-ganciclovir","SBS-RT")])
```


# SBS-RT signatures

## Comparison of novel signatures extracted by the programs

```{r,fig.width=6,fig.height=4}
# Comparison of the novel extracted signature 'SBS-ganciclovir' by hdp and signatureAnalyzer.
hdp_rt1 <- hdp_all$hdp_9
siganal_rt1 <- siganal_all$siganal_7
names(hdp_rt1) <- mut_types$V1
names(siganal_rt1) <- mut_types$V1

pp <- plot_compare_profiles(hdp_rt1,siganal_rt1,condensed = TRUE,profile_ymax=0.1,profile_names = c("hdp_SBS-ganciclovir", "siganal_SBS-ganciclovir"))
pp
pdf("outputs/SBS-ganciclovir_compare_hdp_signatureAnalyzer.pdf",width = 10,height = 4,useDingbats = F)
pp
dev.off()

# Comparison of the novel extracted signature 'SBS-ganciclovir' by hdp and SBSA from Kucab et al.
pp <- plot_compare_profiles(hdp_rt1,SBSA$SBSA,condensed = TRUE,profile_ymax=0.1,profile_names = c("hdp_SBS-ganciclovir", "SBSA"))
pp
pdf("outputs/SBS-ganciclovir_compare_hdp_SBSA.pdf",width = 10,height = 4,useDingbats = F)
pp
dev.off()

# Comparison of the novel extracted signature 'SBS-RT' by hdp and signatureAnalyzer.
hdp_rt2 <- hdp_all$hdp_10
siganal_rt2 <- siganal_all$siganal_8
names(hdp_rt2) <- mut_types$V1
names(siganal_rt2) <- mut_types$V1

pp <- plot_compare_profiles(hdp_rt2,siganal_rt2,condensed = TRUE,profile_ymax=0.1,profile_names = c("hdp_SBS-RT", "siganal_SBS-RT"))
pp
pdf("outputs/SBSRT2_compare_hdp_signatureAnalyzer.pdf",width = 10,height = 4,useDingbats = F)
pp
dev.off()



# Plot SBS-ganciclovir (SBSA) and SBS-RT (unkwown) signatures
p <- plotSignature2(sbsrt$`SBS-ganciclovir`,"SBS-ganciclovir",mut_types$V1)
p
pdf("outputs/SBS-ganciclovir.pdf", height=3, width=10,useDingbats = F)
p
dev.off()
p <- plotSignature2(sbsrt$`SBS-RT`,"SBS-RT",mut_types$V1)
p
pdf("outputs/SBS-RT.pdf", height=3, width=10,useDingbats = F)
p
dev.off()
```

## Further comparison of novel SBS-RT with Kucab et al. and additional comparison plots

```{r}
# Summarize all cosine similarities for SBS-RT and COSMIC
f <- data.frame(matrix(0, ncol=1, nrow = ncol(cancer_signatures)))
colnames(f) <- "SBS-RT"
rownames(f) <- colnames(cancer_signatures)
col_sample <- sbsrt$`SBS-RT`
for (j in 1:ncol(cancer_signatures)) {
    col_cosmic <- cancer_signatures[,j]
    c <- cosineSimilarity(col_cosmic,col_sample)
    f[j,1] <- c
}
f$sig <- rownames(f)

f_result <- f[,c(2,1)]

f$label <- apply(f,1,function(x) ifelse(x["SBS-RT"] > 0.55,x["sig"],""))
g <- ggplot(f,aes(x=as.factor(sig),y=`SBS-RT`))+
  geom_point(color="#73948D") +
  theme_classic() + ylab("Cosine similarity with SBS-RT") + ylim(0,0.7) +
  geom_text_repel(aes(label = label),
                   box.padding   = 0.35,
                   point.padding = 0.5,
                   segment.color = 'grey50',
                   size=3,max.overlaps = 50)+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

# Summarize all cosine similarities for SBS-RT and Kucab et al.
f <- data.frame(matrix(0, ncol=1, nrow = ncol(kucab)))
colnames(f) <- "SBS-RT"
rownames(f) <- colnames(kucab)
col_sample <- sbsrt$`SBS-RT`
for (j in 1:ncol(kucab)) {
    col_cosmic <- kucab[,j]
    c <- cosineSimilarity(col_cosmic,col_sample)
    f[j,1] <- c
}
f$sig <- rownames(f)

f_result <- rbind(f_result,f)

f$label <- apply(f,1,function(x) ifelse(x["SBS-RT"] > 0.45,x["sig"],""))
g2 <- ggplot(f,aes(x=as.factor(sig),y=`SBS-RT`))+
  geom_point(color="#73948D") +
  theme_classic() + ylab("Cosine similarity with SBS-RT") + ylim(0,0.7) +
    geom_text_repel(aes(label = label),
                   box.padding   = 0.35,
                   point.padding = 0.5,
                   segment.color = 'grey50',
                   size=3,max.overlaps = 50)+
  theme(axis.title =element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

pdf("outputs/SBS-RT_cossim_KucabC.pdf", height=3, width=3,useDingbats = F)
g2
dev.off()

grid.arrange(g,g2,ncol=2)
pdf("outputs/SBS-RT_cossim_COSMIC_KucabC.pdf", height=1.5, width=6.75,useDingbats = F)
grid.arrange(g,g2,ncol=2)
dev.off()

write.table(f_result, "outputs/supplementaryTableX_SBS-RT_cossims.tsv", sep="\t", col.names=T, row.names = F, quote=F)

# Decomposition of SBS-RT using COSMIC + Kucab et al.
decompose_into_N_pcawg(hdp[,c("hdp_10"),drop=FALSE],"SBS-RT",cbind(cancer_signatures,kucab),"outputs/SBS-RT_decompose_COSMIC_Kucab")
```

# Clustered mutational signatures

## Load data
```{r}
# Results
hdp <- fread("supplementary_tables/supplementaryTableX_hdp_extracted_signatures_CLUSTERED.tsv",header=TRUE)
sigpro <- fread("supplementary_tables/supplementaryTableX_sigpro_extracted_signatures_CLUSTERED.tsv",header=TRUE)
siganal <- fread("supplementary_tables/supplementaryTableX_siganal_extracted_signatures_CLUSTERED.tsv",header=TRUE)
sigfit <- fread("supplementary_tables/supplementaryTableX_sigfit_extracted_signatures_CLUSTERED.tsv",header=TRUE)

# Signatures present genome-wide + clustered mutational signatures known to be present in CLL and our study
sigs_clust <- c("SBS1", "SBS5", "SBS8", "SBS9", "SBS18", "SBS2", "SBS13", "SBS17b","SBSA","SBS84","SBS85","SBS-ganciclovir","SBS-RT")
```

## Plot clustered extracted signatures 

Plot clustered extracted signatures by each program: HPD, SigProfiler, SignatureAnalyzer and sigfit

```{r}
# SigProfiler
p <- plot_extracted_sigs(sigpro,mut_types$V1,"SigProfiler")
p
pdf("outputs/sigprofiler_extracted_signatures_CLUSTERED.pdf",width = 10,height = 4,useDingbats = F)
p
dev.off()

# HDP
# Remove artefacts
# hdp <- hdp[,!colnames(hdp) %in% artefacts]
p <- plot_extracted_sigs(hdp,mut_types$V1,"HDP")
p
pdf("outputs/hdp_extracted_signatures_CLUSTERED.pdf",width = 10,height = 14,useDingbats = F)
p
dev.off()

# SignatureAnalyzer
p <- plot_extracted_sigs(siganal,mut_types$V1,"SignatureAnalyzer")
p
pdf("outputs/SignatureAnalyzer_extracted_signatures_CLUSTERED.pdf",width = 10,height = 4,useDingbats = F)
p
dev.off()

# sigfit
p <- plot_extracted_sigs(sigfit,mut_types$V1,"sigfit")
p
pdf("outputs/sigfit_extracted_signatures_CLUSTERED.pdf",width = 10,height = 4,useDingbats = F)
p
dev.off()
```

## Assign and decompose clustered extracted signatures


### SignatureAnalyzer

```{r,fig.width=10,fig.height=2}
res <- assign_to_one_pcawg(siganal,cancer_signatures2[,colnames(cancer_signatures2) %in% sigs_clust])
res[1]
res[2]

# We assign:
# siganal_clust1, SBS84, 0.89
# siganal_clust2, SBS85, 0.79
# siganal_clust3, SBS5, 0.82

mod1 <- nnls(as.matrix(cancer_signatures[,c("SBS5","SBS85")]),siganal$siganal_clust2)
mergesig <- mergeSignature2(as.data.frame(cancer_signatures),c("SBS5","SBS85"),mod1$x,mut_types$V1,"siganal_clust2=SBS5+SBS85")
mergesig[[2]]
weights <- paste(round(mod1$x,2),collapse="_")
weights_norm <- paste(round(mod1$x/sum(mod1$x),2),collapse="_")
cossim <- round(cosineSimilarity(mergesig[[1]],siganal$siganal_clust2),2)
print(paste0("SBS5+SBS85 - siganal_clust2 ",weights_norm," ",cossim))
```

### HDP

```{r,fig.width=10,fig.height=2}
res <- assign_to_one_pcawg(hdp,cancer_signatures2[,colnames(cancer_signatures2) %in% sigs_clust])
res[1]
res[2]

# We assign:
# hdp_clust1 - SBS84, 0.91
# hdp_clust2 - SBS85, 0.79

mod1 <- nnls(as.matrix(cancer_signatures[,c("SBS5","SBS85")]),hdp$hdp_clust2)
mergesig <- mergeSignature2(as.data.frame(cancer_signatures),c("SBS5","SBS85"),mod1$x,mut_types$V1,"hdp_clust2=SBS5+SBS85")
mergesig[[2]]
weights <- paste(round(mod1$x,2),collapse="_")
weights_norm <- paste(round(mod1$x/sum(mod1$x),2),collapse="_")
cossim <- round(cosineSimilarity(mergesig[[1]],hdp$hdp_clust2),2)
print(paste0("SBS5+SBS85 - hpd_clust2 ",weights_norm," ",cossim))
```

### SigProfiler

```{r,fig.width=10,fig.height=2}
res <- assign_to_one_pcawg(sigpro,cancer_signatures2[,colnames(cancer_signatures2) %in% sigs_clust])
res[1]
res[2]

# We assign:
# sigpro_clust1, SBS84, 0.84
# sigpro_clust2, SBS85, 0.81
# sigpro_clust3, SBS5, 0.85

mod1 <- nnls(as.matrix(cancer_signatures[,c("SBS5","SBS85")]),sigpro$sigpro_clust2)
mergesig <- mergeSignature2(as.data.frame(cancer_signatures),c("SBS5","SBS85"),mod1$x,mut_types$V1,"sigpro_clust2=SBS5+SBS85")
mergesig[[2]]
weights <- paste(round(mod1$x,2),collapse="_")
weights_norm <- paste(round(mod1$x/sum(mod1$x),2),collapse="_")
cossim <- round(cosineSimilarity(mergesig[[1]],sigpro$sigpro_clust2),2)
print(paste0("SBS5+SBS85 - sigpro_clust2 ",weights_norm," ",cossim))
```

### sigfit

```{r,fig.width=10,fig.height=2}
res <- assign_to_one_pcawg(sigfit,cancer_signatures2[,colnames(cancer_signatures2) %in% sigs_clust])
res[1]
res[2]

# We assign:
# sigfit_clust2 - SBS84, 0.84
# sigfit_clust3 - SBS85, 0.81
# sigfit_clust1 - SBS5, 0.87

mod1 <- nnls(as.matrix(cancer_signatures[,c("SBS5","SBS85")]),sigfit$sigfit_clust2)
mergesig <- mergeSignature2(as.data.frame(cancer_signatures),c("SBS5","SBS85"),mod1$x,mut_types$V1,"sigfit_clust2=SBS5+SBS85")
mergesig[[2]]
weights <- paste(round(mod1$x,2),collapse="_")
weights_norm <- paste(round(mod1$x/sum(mod1$x),2),collapse="_")
cossim <- round(cosineSimilarity(mergesig[[1]],sigfit$sigfit_clust2),2)
print(paste0("SBS5+SBS85 - sigpro_clust2 ",weights_norm," ",cossim))
```




# Session 
```{r}
sessionInfo()
```